How to publish in Biological Sciences: a guide for the uninitiated

### 2021-12-15 2.03-suggestreviewers.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

You will often be asked to suggest reviewers as you submit your manuscript. This can be part of your [cover letter](letter.html), or data that you need to fill in during the submission process. Keep in mind that supplying a name alone will not be considered sufficient, you will need their name, the name of their institution and their email address. Some journals will not accept addresses that do not come from institutions unless they can be verified. The reason for this is that there has been an outbreak of fraudulent reviewers, where authors suggested fake reviewers or gave email addresses that they manufactured [see @brainard2018what].

Following the [spirit of peer review](#spirit), you should suggest people that you think will provide insightful reviews. These could be people that work in the same area (but that aren't involved in your study). Personally, I would be likely to name the same set of people that I would invite myself if I were editing. People that I believe would provide a constructive and unbiased review.

@@

As an editor, I will often use one of the reviewers suggested by the authors if they fit my (following) criteria. I will make sure that I balance this with another reviewer or two not suggested by the authors. To select reviewers, I will [read the submission](editors.html) and look for citations of similar studies, or techniques (depending on the type of paper). From the citation, I will find the paper (preferably published in the last 3 years) and use the corresponding author's address. I will also try to visit the website of the senior (last) author to see whether they have a lab that is still active in this area, and especially I am looking for post-docs or Early Career Scientists who cover the same topic. If they are there, I will invite them. If not, I will write to the lab head in the hope that they may well recommend one of their post-docs. This process can take a long time, especially hunting down email addresses that constantly bounce back from the journal's editorial manager software.

@@

If the manuscript is a resubmission, I will try to use the same reviewers that made previous reviews. This isn't always possible, and I know that it is a source of upset for authors when they receive new reviews on a second or even third round of review. I don't think that any editors will do this deliberately, but reviewers do have the option of indicating that they are not willing to read a resubmission. If this is the case for most or all of the reviewers, then you are likely to face an entirely fresh round of peer review.

>>>>

You will often be asked to suggest reviewers as you submit your manuscript. This can be part of your [cover letter](#letter), or data that you need to fill in during the submission process. Keep in mind that supplying a name alone will not be considered sufficient, you will need their name, the name of their institution and their email address. Some journals will not accept addresses that do not come from institutions unless they can be verified. The reason for this is that there has been an outbreak of fraudulent reviewers, where authors suggested fake reviewers or gave email addresses that they manufactured [see @brainard2018what].

In keeping with the [spirit of peer review](#spirit), you should suggest people that you think will provide insightful reviews. These could be people that work in the same area (but that aren't involved in your study). Personally, I would be likely to name the same set of people that I would invite myself if I were editing. People that I believe would provide a constructive and unbiased review.

@@

As an editor, I will often use one of the reviewers suggested by the authors if they fit my (following) criteria. I will make sure that I balance this with another reviewer or two not suggested by the authors. To select reviewers, I will [read the submission](#editors) and look for citations of similar studies, or techniques (depending on the type of paper). From the citation, I will find the paper (preferably published in the last 3 years) and use the corresponding author's address. I will also try to visit the website of the senior (last) author to see whether they have a lab that is still active in this area, and especially I will be looking for post-docs or Early Career Scientists who cover the same topic. If they are there, I will invite them. If not, I will write to the lab head in the hope that they may well recommend one of their post-docs. This process can take a long time, especially [hunting down email addresses](#website) that constantly bounce back from the journal's editorial manager software.

@@

If the manuscript is a resubmission, I will try to use the same reviewers that made previous reviews. This isn't always possible, and I know that it is a source of upset for authors when they receive new reviews on a second or even third round of review. I don't think that any editors will do this deliberately, but reviewers do have the option of indicating that they are not willing to reread a resubmission. If this is the case for most or all of the reviewers, then you are likely to face an entirely fresh round of peer review.

### 2021-12-15 2.04-whichjournal.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

In this section I provide steps to go from a potentially very long list of journals where you might publish, to choosing the one where you will submit your manuscript first. As you work through the steps, keep notes on why you exclude journals at each step. You may want to revisit these criteria later, or explain to your coauthors why a particular journal was not considered.

@@

- The editor will [desk reject](#desk) your manuscript if it is not appropriate (and you'll waste their time)
- Many reviewers give feedback tailored to the journal, and this will not be appropriate in the wrong journal
- Perhaps most importantly, you are likely to get additional readers in the correct journal that might not find your paper otherwise
- Note that there are no-Impact Factor journals for which this above is not relevant, and these are [discussed separately](#noif)

This section is geared toward the 'current' model in the life sciences of journal title as a measure of quality. The reality is that this model is shifting, and we may well end up with another (better) [open model](#closedtoopen) in the future based around [preprint servers](#preprints2) and [overlay journals](#overlay). The current reliance on journal titles is driving a [toxic culture](#PruittData) in biological sciences that is becoming increasingly recognised, even by those who are [gatekeepers](#gatekeepers) for this model (see [Part IV](#retract)). To reflect this viewpoint, and help drive a culture of change, I have placed [Impact Factor](#impactfactor) outside of this list of steps. In doing so, I acknowledge that you and your co-authors may well be operating inside the current publisher driven model advocating for Impact Factor as a step with high priority. Even if this is your reality, I hope that you will acknowledge and appreciate that this is not the way that journals should be chosen.

@@

Journals have different hierarchies of scope. Some journals attempt to take on the full gambit of science (e.g. Science and Nature) while others are only interested in a particular taxonomic group. In general, the impact factor of the journal is likely to be linked to the diversity of the scope. This is not always the case. There are some taxonomically specific journals with high impact factors, and there are some general journals with low impact factors.

Once you have determined your subject area make a list of the journals within this area. Based on your reading of literature pertinent to your manuscript, try to decide whether your manuscript is likely to be accepted by a journal with a high ranking impact factor, a medium ranking journal, or a low-ranking journal. Note that in your discipline it may be more appropriate to use a five or ten year impact factor to rank your journals, instead of the normal two. Sort the list of journals within the subject area by impact factor and then select the ones that are either high (>10), medium (>4), or low (<4).

### The Journal Scope {#scope}

Every journal has a scope that is stated on their website. Your manuscript must fit into the scope of the journal that you select. In order to get your candidate list of journals you now need to visit their websites to look at their scope in detail. Ill fitting scope is the most common cause of a [desk rejection](#desk) [@schimel2014editors; @teixeiradasilva2018establishing].

@@

Academic publishing started with society journals, and I think that they are still worth supporting if you can. Your mentor or co-authors may be members of particular academic societies, and may have a preference therefore to publish in their society's journal. See below for other potential advantages in publishing in a society journal.

@@

### Is there a specail issue?
Special issues are a great way for your work to get better exposure, both to others that are participating in the special issue, and to those who come across the entire issue later (see [Part I](#citations)).

@@

In an ideal world, there wouldn’t be any more barriers to you publishing your contribution to the collective of scientific knowledge. Not only is it not an ideal world, but I would argue that there has never been a less ideal time for publishing science. Greedy publishers have taken publicly funded science and made the public pay for it time and again. Read more on this [later in the book](#paywall).

@@

Some journals are exclusively [Open Access (OA)](#openaccess1) meaning that you will need to pay (unless they are [Diamond OA](#diamond)) in order to publish your accepted manuscript (author pays). Different types of OA are covered in a subsequent chapter (see [Part II](#openaccess1)). More information on OA is provided in [Part IV](#openaccess), and I encourage you to read this before you decide to pay any research money toward OA publishing Article Processing Charges (APCs).

@@

### No impact factor journal #NOif

There's no problem to publish in a journal that has no interest in impact factor. Indeed, it can be seen as 'the right thing to do', with respect to confirmation bias. However, if you are submitting to such a journal, then you should be aware that this is what their speciality is: an emphasis on the technical soundness, rather than whether or not a significant result was found. The list of 'no impact factor journals' is growing, and they tend to be online only, open access and carry substantial article processing charges.

## Shortlist
Once you have your shortlist of journals to consider, take it to your mentor. Together with your coauthors rework your list into something that you both all agree with. Rank your list by journals that you want to try first and those that are your last options at the end.

@@

>>>>

In this section I provide steps to go from a potentially very long list of journals where you might publish, to choosing the one where you will submit your manuscript first. As you work through the steps, keep notes on why you exclude journals at each step. You may want to revisit these criteria later, or explain to your co-authors why a particular journal was not considered.

@@

- The editor will [desk reject](#desk) your manuscript if it is not appropriate (and you'll waste their time).
- Many reviewers give feedback tailored to the journal, and this will not be appropriate in the wrong journal.
- Perhaps most importantly, you are likely to get additional readers in the correct journal that might not find your paper otherwise.
- Note that there are no-Impact Factor journals for which this above is not relevant, and these are [discussed separately](#noif).

This section is geared toward the 'current' model in the life sciences of journal title as a measure of quality. The reality is that this model is shifting, and we may well end up with another (better) [open model](#closedtoopen) in the future based around [preprint servers](#preprints2) and [overlay journals](#overlay). The current reliance on journal titles is driving a [toxic culture](#PruittData) in biological sciences that is becoming increasingly recognised, even by those who are [gatekeepers](#gatekeepers) for this model (see [Part IV](#retract)). To reflect this viewpoint, and help drive a culture of change, I have placed [Impact Factor](#impactfactor) outside of this list of steps. In doing so, I acknowledge that you and your co-authors may well be operating inside the current publisher driven model advocating for Impact Factor as a step with high priority [see @niles2020why for perceived importance of different factors in choosing a journal]. Even if this is your reality, I hope that you will acknowledge and appreciate that this is not the way that journals should be chosen.

As an ECR, you are likely to need to think of review, promotion and [tenure](#tenure) when considering where to publish. An important point to be aware of is how your institution perceives quality, prestige and impact. These are rather nebulous terms that can sometimes be well defined by institutions, but that faculty often mix-up these terms and frequently rely on [Impact Factor](#impactfactor) to define them [@morales2021how]. You should, therefore, consult documentation on review, promotion and [tenure](#tenure) at your institution, and/or consult your [mentor](#mentor). You may then want to add these definitions at the top of your table in your journal selection process.

@@

Journals have different hierarchies of scope. Some journals attempt to take on the full gambit of science (e.g. *Science* and *Nature*) while others are only interested in a particular taxonomic group. In general, the impact factor of the journal is likely to be linked to the diversity of the scope. This is not always the case. There are some taxonomically specific journals with high impact factors, and there are some general journals with low impact factors.

Once you have determined your subject area make a list of the journals within this area. Based on your reading of literature pertinent to your manuscript, try to decide whether your manuscript is likely to be accepted by a journal with a high rank. Many databases rank journals into quartiles (Q1 to Q4) based on their subject area. For example, you might see that a journal is ranked Q1 for Biodiversity Conservation, but Q2 for Ecology. These groupings will change depending on the database. These quartiles can be useful groups to set your aims at, but make sure that they correspond with the subject area chosen for your submission.

### The Journal Scope and Readership {#scope}

Every journal has a scope that is stated on their website. Your manuscript must fit into the scope of the journal that you select. In order to get your candidate list of journals you now need to visit their websites to look at their scope in detail. Ill fitting scope is the most common cause of a [desk rejection](#desk) [@schimel2014editors; @teixeiradasilva2018establishing]. Journal Scope is also most strongly aligned to readership, that is the specific readership that you want to reach when publishing your article. Readership was recognised as the strongest factor for North American faculty in selecting a journal [@niles2020why].

@@

Academic publishing started with society journals, and I think that they are still worth supporting if you can. You or your co-authors may be members of particular academic societies, and may have a preference therefore to publish in their society's journal. See below for other potential advantages in publishing in a society journal.

@@

### Is there a special issue?
Special issues are a great way for your work to get better exposure, both to others that are participating in the special issue, and to those who come across the entire issue later (see [Part I](#citations)).

@@

In an ideal world, there wouldn’t be any more barriers to you publishing your contribution to the collective of scientific knowledge. Not only is it not an ideal world, but I would argue that there has never been a less ideal time for publishing science. Greedy publishers have taken publicly funded science and made the content creators pay for it time and again. Read more on this [later in the book](#paywall).

@@

Some journals are exclusively [Open Access (OA)](#openaccess1) meaning that you will need to pay (unless they are [Diamond OA](#diamond)) in order to publish your accepted manuscript (author pays). Different types of OA are covered in a subsequent chapter (see [Part II](#openaccess1)). More information on OA is provided in [Part IV](#openaccess), and I encourage you to read this before you decide to pay any research money toward OA publishing [Article Processing Charges](#APC) (APCs).

@@

Be aware that there are trade-offs when deciding whether or not to submit your manuscript to a journal with high or low ranking Impact Factor [@tiokhin2021honest]. By continuing to support publishers and their attempts to push academics into journals with different ranks, we suffer an overall loss in our attempts to communicate our results.

### No impact factor journal {#NOif}

In this book I refer to journals that have no interest in their [Impact Factor](#impactfactor) as 'no impact factor'. Of course, if these journals are listed by impact factor providing databases (and most are), then they will have an impact factor and this can be high - even higher than other journals that attempt to inflate their impact factors (see [Part IV](#PLOS)). Instead, what I mean is that these journals are operating at minimum standards of technical soundness (so called '[negative selection](#AKP)'), instead of the Anna Karenina Principle (AKP) selection: only papers that are thought by their gatekeepers to uphold or increase the journal IF.

There's no problem to publish in a journal that has no interest in Impact Factor. Indeed, it can be seen as 'the right thing to do', with respect to confirmation bias. This is the way that all science was published prior to the dominance of [citation metrics](#citations). If you are submitting to a 'no impact factor journal' then you should be aware that this is what their speciality is: an emphasis on the technical soundness, rather than whether or not a significant result was found. You need to be aware as this is how inclusion of such publications in your CV will be regarded by peers. The list of 'no impact factor journals' is growing, and they tend to be online only, [Gold OA](#gold) and carry substantial [article processing charges](#gold).

## Shortlist
Once you have your shortlist of journals to consider, take it to your mentor. Together with your co-authors rework your list into something that you all agree with. Rank your list by journals that you want to try first and those that are your last options at the end.

@@

It seems like a lot of trouble to go to, but really making such a list helps you rationalise your choices and will save you a lot of time if you end up submitting to lots of different journals. Keep the list because your next publication (or another not to long down the road) might also be in the same area.

### 2021-12-15 2.05-openaccess.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

(ref:Paywall-graphic) **The world of Open Access (OA) is built on [shaky foundations](#openaccess).** Much more of journal content is still within the shadow of the paywall, even for journals that practice Hybrid and Bronze OA. When you are inside your institution, you may not notice the paywall due to seamless integration with many literature databases. You will be far better acquainted with the paywall if you are from a smaller institution or trying to access literature from outside your institution. Once you are the wrong side of the paywall, your options for most content will be Grey and Black OA. There are regular assaults on these two initiatives by publishers.

@@

## Gratis or Libre OA {#libre}
Some definitions of OA extend beyond simply being allowed to read the text (Gratis OA) to it bearing a “CC-BY” Creative Commons license. Libre OA means that in addition to being able to share, redistribute or copy the content, the CC-BY (4.0) licence allows others to “remix, transform, and build upon the material for any purpose, even commercially”.

A good example of this is my ability to publish the following Figure \@ref(fig:Piwowar) from Piwowar et al. [-@piwowar2018state] as the content of _PeerJ_ is published under CC-BY licence.

@@

## Green OA {#green}
This is an option no matter what journal you publish in, including [closed journals](#closed) behind the paywall. The accepted manuscript, before it has been typeset, is deposited in your institutional repository; usually your library hosts this. This is sometimes referred to as a **postprint** (manuscript with changes made following peer review) - the same version that has been accepted by the journal. Once your paper is published by the journal anyone can access the manuscript that was accepted because it is available free from your institutional repository. However the typeset paper will remain behind the publishers’ paywall (i.e. [closed](#closed)). Note that this version will not be the [Version of Record](#VoR), it may be that some of the text will change with [copy editing](#proofs). However, for people who want Open Access your work, this Green OA version should suffice.

When you deposit your author version to your institutional repository, make sure that you add a [Creative Commons Attribution](https://creativecommons.org/) license (CC BY). In theory, this turns your work into something that is [Libre](#libre) Open Access. Although the wording will be the same in the [Closed OA](#closed) version eventually available from the publisher, the prior attribution of CC BY to your Green OA work should mean that all subsequent versions are Libre OA. There are a number of reasons why this may not be the case, and you should be aware of two of these [@khoo2021why]. First are your obligations to your funders. For example, your funders may have specific requirements about how you publish, and prior rights attribution may conflict. Funders may have other restrictions on your publishing including embargo periods. With many publishers you may be required to sign an acceptance letter that specifically states that it's supersedes any previous licensing commitments including self archiving copies that you make CC BY. Your previous CC BY commitments to your pre-print may then place you in a legal conflict. Publishers are unlikely to assert their license unless you try to assert yours.

If you are unsure about whether or not to deposit a copy in your institutional repository and what license should be attributed to this, then ask you a librarian. You may need to finish them with details of your funding and potential target journals where you'd like to publish.

## Bronze OA
Bronze Open Access is an option at the discretion of the publishers. Only some publishers do this for some of their content. For example, they may decide that a certain thematic issue should be open access, an editorial, or review article. There is some debate about whether this bronze model is truly OA as it often still carries copyright restrictions.

## Delayed OA
Some journals choose to have their archived content available for all readers without a paywall. This could be after two or five years. Like Bronze OA, there is some debate about whether delayed OA is truly open as it often still carries the publishers’ copyright restrictions.

## Gold OA {#gold}
In a journal with gold open access it is compulsory for you to pay the open access fee (Article Processing Charge) in order to publish your paper. These fees can be very large, so if your choice of journal is gold open access then make sure you know where the fees are coming from. The history of Gold OA is an interesting one that you can read in [Part IV](#PLOS).

Examples of gold open access journals are [PLoS one](https://journals.plos.org/plosone/) and [PeerJ](www.peerj.com), and an increasing hoard of [no impact factor journals](#noif). The advantage to these journals is that as soon as you publish your work in them, everyone will have access to it without any paywall. The disadvantage is that you may not be able to afford to publish there. The disadvantages of these journals are now becoming very prominent in the Biological Sciences, such that you may be excluded from increasing numbers of journals in your field. Alarmingly, journals with higher impact factors are also charging ever increasing sums to publish Gold OA [@gray2020sorry]. Gray [-@gray2020sorry] makes the important point that prestige (often confused with [Impact Factor](#impactfactor)) is being allocated a higher price in Gold OA, that is likely to disadvantage and disenfranchise scientists from less wealthy institutions and countries. This is likely to reinforce an increasing dichotomy between rich and poor researchers. We will take another look at who pays for OA [later in the book](#openaccess).

Watch out for predatory journals among those journals with Gold OA. Predatory publication will be [discussed in full later](#predatory)).

## Hybrid OA
Hybrid OA journals are increasingly the norm. You can decide upon acceptance of your manuscript whether or not you want to pay the fee to make your article open access. Again, note that your institution may have a deal with the publisher that means that anything you publish there is open access. It is well worth knowing these things in advance before you submit. If you can't afford to pay the open access fee then your manuscript will remain behind a paywall and be only available to subscribers.

A slight variation on hybrid OA is when you are a member of an academic society that allows its members to publish open access without extra payment. As a student you are likely to get very discounted membership to an academic society Which might make it very cheap for you to publish open access with them.

## Platinum or Diamond OA {#diamond}
This is without doubt the best OA model, and the one that we should all strive for. In a platinum or diamond open access journal you do not have to pay any money but everything that is published is open access. In order to do this these journals are often subsidised by governmental or philanthropic agencies. Some university presses are also in the habit of publishing platinum or diamond OA when it meets with their stated mission. These journals are very rare but they do exist, and it is well worth looking for them.

@@

## Black OA
This refers to the placement of published material onto a pirate website such as [Sci-Hub](https://sci-hub.se/). Sci-Hub is considered by most governments to be illegal and may be blocked by your institution or country. However they do a great job in making scientific publications open access and need to be supported for that. Many scientists all over the world depend on Sci-Hub in order to access literature and therefore conduct research. In addition, there are a number of other Black OA sites: [libgen](http://libgen.li/),  [Unpaywall.org](https://unpaywall.org/), [Open Access Button](https://openaccessbutton.org/), [freefullpdf](http://www.freefullpdf.com/).

## Grey OA
You can find grey OA repositories of published material on Academic Social Network sites in which you need membership to access such as [ResearchGate](https://www.researchgate.net/) and [Academia.edu](https://www.academia.edu/). The legality of such sites is regularly questioned [see @piwowar2018state for more details]. There has been legal action with thousands of members being issues with take-down notices.

@@

## Unsure what you can legally do with a published paper?
If you don’t know what level of copyright exists on something that you have published, then you can find an aggregated set of publisher policies at [Sherpa Romeo](https://v2.sherpa.ac.uk/romeo/). This is a really nice database which provides a very simple summary by journal. You can also use this to check out a journal that you are thinking of publishing with.

>>>>

(ref:Paywall-graphic) **The world of Open Access (OA) is built on [shaky foundations](#openaccess).** Much more of journal content is still within the shadow of the paywall, even for journals that practice Hybrid and [Bronze](#bronze) OA. When you are inside your institution, you may not notice the paywall due to seamless integration with many literature databases. You will be far better acquainted with the paywall if you are from a smaller institution or trying to access literature from outside your institution. Once you are the wrong side of the paywall, your options for most content will be Grey and Black OA. There are regular assaults on these two initiatives by publishers.

@@

##Open Access 
There is a bewildering array of different Open Access types. Entries in the list below are not mutually exclusive. Moreover, you may not be aware how your manuscript may be treated, and it is not always possible to predict the level of OA your article will be published with in many journals. Each journal will likely have their own policy, tied to the publisher in many cases.

### Gratis OA {#gratis}
Gratis OA simply refers to the ability for anyone to access the text free, and without any [paywall](#paywall).

### Libre OA {#libre}
Some definitions of OA extend beyond simply being allowed to access the text for free ([Gratis OA](#gratis)) to it bearing a license so that you can share, redistribute and reuse (Libre OA). With Libre OA everyone is able to share, redistribute or copy the content. The Creative Commons (CC-BY 4.0) licence allows others to “remix, transform, and build upon the material for any purpose, even commercially”.

A good example of this is my ability to publish the following Figure \@ref(fig:Piwowar) from Piwowar et al. [-@piwowar2018state] as the content of _PeerJ_ is published under CC BY licence.

@@

### Green OA {#green}
This is an option no matter what journal you publish in, including [closed journals](#closed) behind the paywall. The accepted manuscript, before it has been typeset, is deposited in your institutional repository; usually your library hosts this. Alternatively, you can post your accepted manuscript onto another platform such as [ShareYourPaper](Shareyourpaper.org), which is a really easy and simple way to make sure that your content becomes universally available via tools like [Unpaywall](unpaywall.org). This is sometimes referred to as a **postprint** (manuscript with changes made following peer review) - the same version that has been accepted by the journal. Once your paper is published by the journal anyone can access the manuscript that was accepted because it is available free from your institutional repository. However the typeset paper will remain behind the publishers’ paywall (i.e. [closed](#closed)). Note that this version will not be the [Version of Record](#VoR), it may be that some of the text will change with [copy editing](#proofs). However, for people who want Open Access to your work, this Green OA version should suffice.

When you deposit your author version to your institutional repository, make sure that you add a [Creative Commons Attribution](https://creativecommons.org/) license (CC BY). In theory, this turns your work into something that is [Libre](#libre) Open Access [see @eglen2021primer]. Although the wording will be the same in the [Closed OA](#closed) version eventually available from the publisher, the prior attribution of CC BY to your Green OA work should mean that all subsequent versions are [Libre](#libre) OA. There are a number of reasons why this may not be the case, and you should be aware of two of these [@khoo2021why]. First are your obligations to your funders. For example, your funders may have specific requirements about how you publish, and prior rights attribution may conflict. Funders may have other restrictions on your publishing including embargo periods. With many publishers you may be required to sign an acceptance letter that specifically states that it's supersedes any previous licensing commitments including self archiving copies that you make CC BY. Your previous CC BY commitments to your pre-print may then place you in a legal conflict. Publishers are unlikely to assert their license unless you try to assert yours.

If you are unsure about whether or not to deposit a copy in your institutional repository and what license should be attributed to this, then ask your librarian. You may need to furnish them with details of your funding and the journal where your manuscript has been accepted for publication.

### Bronze OA {#bronze}
Bronze Open Access is an option at the discretion of the publishers. Only some publishers do this for some of their content. For example, they may decide that a certain thematic issue should be open access, an editorial, or review article. There is some debate about whether this bronze model is truly OA as it often still carries copyright restrictions.

### Delayed OA {#delayed}
Some journals choose to have their archived content available for all readers without a paywall. This could be after two or five years. Like [Bronze OA](#bronze), there is some debate about whether delayed OA is truly open as it often still carries the publishers’ copyright restrictions.

### Gold OA {#gold}
In a journal with gold open access it is compulsory for you to pay the open access fee (or Article Processing Charge, often referred to as [APC](#APC)) in order to publish your paper. These [APC](#APC)s can be very large (see [below](#APC)), often more than \$1000 (see [Part IV](#money)). If your choice of journal is gold open access then make sure you know where the fees are coming from. The history of Gold OA is an interesting one that you can read in [Part IV](#PLOS).

Examples of gold open access journals are _PLOS ONE_, _F1000Research_ and _PeerJ_, and an increasing hoard of [no impact factor journals](#noIF). The advantage to these journals is that as soon as you publish your work in them, everyone will have access to it without any paywall. The disadvantage is that you may not be able to afford to publish there, and their costs are usually inflated to maximise profit [@grossmann2021current]. The disadvantages of these journals are now becoming very prominent in the Biological Sciences, such that you may be excluded from increasing numbers of journals in your field. Alarmingly, journals with higher impact factors are also charging ever increasing [APC](#APC)s to publish Gold OA [@gray2020sorry]. Gray [-@gray2020sorry] makes the important point that prestige (often confused with [Impact Factor](#impactfactor)) is being allocated a higher price in Gold OA, that is likely to disadvantage and disenfranchise scientists from less wealthy institutions and countries. This is likely to reinforce an increasing dichotomy between rich and poor researchers. We will take another look at who pays for OA [later in the book](#openaccess).

Watch out for predatory journals among those journals with Gold OA. Predatory publication will be [discussed in full later](#predatory).

### Hybrid OA {#hybrid}
Hybrid OA journals are increasingly the norm. You can decide upon acceptance of your manuscript whether or not you want to pay the fee to make your article open access. Again, note that your institution may have a deal with the publisher that means that anything you publish there is open access. It is well worth knowing these things in advance before you submit. If you can't afford to pay the open access fee then your manuscript will remain behind a paywall and be only available to subscribers.

Note that many journals refer to paying for OA inside a hybrid journal as [Gold OA](#gold). Although this may appear confusing, essentially they are offering the same service as for Gold OA journals but in a hybrid format. A study comparing citation rates for papers published with Hybrid OA, suggests that the ~30% increased citations achieved is equivalent to the same citation increase obtained for making manuscripts available via [Green OA](#green) [@piwowar2018state]. This should be an extra incentive to make an article available via an institutional repository, instead of paying a publisher for Hybrid OA.

A slight variation on hybrid OA is when you are a member of an academic society that allows its members to publish open access without extra payment. As a student you are likely to get very discounted membership to an academic society, which might make it very cheap for you to publish open access with them.

### Platinum or Diamond OA {#diamond}
This is without doubt the best OA model, and the one that we should all strive for. In a platinum or diamond open access journal you do not have to pay any money but everything that is published is open access. Moreover, the content is free to be disseminated (CC BY), but does not allow any for-profit re-use [@fuchs2013diamond]. In order to do this these journals are often subsidised by governmental or philanthropic agencies. Some university presses are also in the habit of publishing platinum or diamond OA when it meets with their stated mission. These journals are very rare but they do exist, and it is well worth looking for them.

@@

### Black OA {#black}
This refers to the placement of published material onto a pirate website such as [Sci-Hub](https://sci-hub.se/). Sci-Hub is considered by most governments to be illegal and may be blocked by your institution or country. However they do a great job in making scientific publications open access and need to be supported for that. Many scientists all over the world depend on Sci-Hub in order to access literature and therefore conduct research. In addition, there are a number of other Black OA sites: [libgen](http://libgen.li/), [freefullpdf](http://www.freefullpdf.com/).

### Grey OA {#grey}
You can find grey OA repositories of published material on Academic Social Network sites in which you need membership to access such as [ResearchGate](https://www.researchgate.net/) and [Academia.edu](https://www.academia.edu/). The legality of such sites is regularly questioned [see @piwowar2018state for more details]. There has been legal action with thousands of members being issues with take-down notices.

## Article Processing Charge - APC {#APC}
The Article Processing Charge, or APC as it's commonly referred to, is the way in which Gold OA journals pay for their services. This is probably one of the most important numbers to look at when considering whether or not to submit your manuscript to Gold OA journals. There are some Gold OA journals that charge very small APCs, literally what it costs them to typeset and host the [VoR](#VoR). But these are not the norm. You will need to read later in the book to find out [what publishers do](#deals) with the [money](#money) you give them, and why having you [believe in their metrics](#impactfactor) helps them take more cash from you and your institution.

@@

## Unsure what you can legally do with a published paper?
If you don’t know what level of copyright exists on something that you have published, then you can find an aggregated set of publisher policies at [Sherpa Romeo](https://v2.sherpa.ac.uk/romeo/). This is a really nice database which provides a very simple summary by journal. You can also use this to check out a journal that you are thinking of publishing with. If you are still in doubt, then consult your librarian.
